Multi-level modeling with Openflexo/\FML

### 2022-02-23 10-conclusions.tex
style guidelines: quotation marks -> \enquote
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. This infrastructure helped us to overcome the limitations of the strict modeling paradigm mentioned in Section \ref{sec:introduction}, this is, the lack of support for different forms of classifications and for the duality type/object. In that sense, our solution seamlessly integrates the type-object pattern which allows us to dynamically create different \emph{instances} of model elements (e.g., \emph{TaskType}, among others) and use them for typing other model elements. This, together with the \FML core support for (level agnostic) classical linguistic instantiation, permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.

The features of \FML enable a great modeling flexibility which, coupled with an appropriate methodology, allow us to achieve 
the multi-level capabilities without dedicated tooling, in order to solve problems which require it. We use virtual models to "implement" levels on demand.
%\noteSM{Say that the flexibility of our approach coupled with a methodology allows us to achieve multi/level capabilities without specific tooling in order to solve problems which require it.}

>>>>

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. This infrastructure
helped us to overcome the limitations of the strict modeling paradigm mentioned
in Sect.~\ref{sec:introduction}, this is, the lack of support for different
forms of classifications and for the duality type/object. In that sense, our
solution seamlessly integrates the type-object pattern which allows us to
dynamically create different \emph{instances} of model elements (\eg, \emph{TaskType}, among others) and use them for typing other model elements. This, together with the \FML core support for (level agnostic) classical linguistic instantiation, permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.

The features of \FML enable a great modeling flexibility which, coupled with an appropriate methodology, allow us to achieve 
the multi-level capabilities without dedicated tooling, in order to solve
problems which require it. We use virtual models to \enquote{implement} levels on demand.

%\noteSM{Say that the flexibility of our approach coupled with a methodology allows us to achieve multi/level capabilities without specific tooling in order to solve problems which require it.}

### 2022-02-23 3-introduction.tex
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

The rest of the paper is organized as follows. Section~\ref{sec:technology} presents our approach and the technology it relies on. Section~\ref{sec:analysis} is the analysis of the problem. We detail our model in section~\ref{sec:model} and show how it satisfies the challenge requirements in section~\ref{sec:requirements}. We assess the modeling solution in section~\ref{sec:discussion}. Section~\ref{sec:relatedwork} presents the related work before the conclusions.%~\ref{sec:conclusions}.

>>>>

The rest of the paper is organized as follows. Sect.~\ref{sec:technology}
presents our approach and the technology it relies on. Sect.~\ref{sec:analysis}
is the analysis of the problem. We detail our model in Sect.~\ref{sec:model}
and show how it satisfies the challenge requirements in
Sect.~\ref{sec:requirements}. We assess the modeling solution in
Sect.~\ref{sec:discussion}. Sect.~\ref{sec:relatedwork} presents the related work before the conclusions.%~\ref{sec:conclusions}.

### 2022-02-23 4-technology.tex
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

This framework relies on the architecture of figure~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Federation Modeling
Language (\FML). Each federated model is an autonomous
element that may evolve with its own tooling. The virtual models
serve as control elements binding the federated models together.
In this paper, we mainly use conceptual models and more precisely \FML its domain specific modeling language. The only use of the federation aspect is on the tooling made for the process language.

A \UML like representation of a simplified metamodel of \FML is provided in figure~\ref{fig:mm}. \FML is a language designed to define models as virtual models. A \emph{virtual model} is composed of a set of \emph{concepts}, while itself being a \emph{concept}. Hence, \emph{virtual models} are structuring units forming architectures while \emph{concepts} are the core entities. A concept has a set of \emph{properties} and \emph{behaviors}. \emph{Properties} can be either simple variables, roles (pointers to a modeling element in a \emph{virtual model} or an external technology-specific model), or properties bound to complex control graphs. \FML is designed to define not only the structure of virtual models but also the collection of actions that can be performed on them. These actions are called \emph{behaviors} and rely on behavioral primitives called \textsf{EditionAction}. These \emph{behaviors} can either be called or triggered by events. The reactive behaviors are useful when a federated model evolution needs to trigger a computation. Note that we do not exploit this possibility in the challenge.

@@

To illustrate this, figure~\ref{fig:BPMNSubsetExample} shows how to realize a small subset of BPMN with Openflexo/FML. Concretely, we have designed a \textit{virtual model} (\texttt{SimplifiedBPMN}) with 3 \textit{concepts} representing a process and its contained tasks. An instance of this \textit{virtual model} is presented in the bottom of the figure, showing instances with their \textsf{instanceOf} relationships with their model.
%
The following listing shows the \FML code of this \textit{virtual model}. Note that the composition relation of figure~\ref{fig:BPMNSubsetExample} between \texttt{BPMNProcess} and \texttt{BPMNTask} is realized in the FML code by concept containment. Whenever an instance of \texttt{BPMNTask} is created, it must be in a container (instance of \texttt{BPMNProcess}) and the two instances get linked together. %\noteFabien{A valider}

>>>>

This framework relies on the architecture of Fig.~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Federation Modeling
Language (\FML). Each federated model is an autonomous
element that may evolve with its own tooling. The virtual models
serve as control elements binding the federated models together.
In this paper, we mainly use conceptual models and more precisely \FML its domain specific modeling language. The only use of the federation aspect is on the tooling made for the process language.

A \UML like representation of a simplified metamodel of \FML is provided in Fig.~\ref{fig:mm}. \FML is a language designed to define models as virtual models. A \emph{virtual model} is composed of a set of \emph{concepts}, while itself being a \emph{concept}. Hence, \emph{virtual models} are structuring units forming architectures while \emph{concepts} are the core entities. A concept has a set of \emph{properties} and \emph{behaviors}. \emph{Properties} can be either simple variables, roles (pointers to a modeling element in a \emph{virtual model} or an external technology-specific model), or properties bound to complex control graphs. \FML is designed to define not only the structure of virtual models but also the collection of actions that can be performed on them. These actions are called \emph{behaviors} and rely on behavioral primitives called \textsf{EditionAction}. These \emph{behaviors} can either be called or triggered by events. The reactive behaviors are useful when a federated model evolution needs to trigger a computation. Note that we do not exploit this possibility in the challenge.

@@

To illustrate this, Fig.~\ref{fig:BPMNSubsetExample} shows how to realize a small subset of BPMN with Openflexo/FML. Concretely, we have designed a \textit{virtual model} (\texttt{SimplifiedBPMN}) with 3 \textit{concepts} representing a process and its contained tasks. An instance of this \textit{virtual model} is presented in the bottom of the figure, showing instances with their \textsf{instanceOf} relationships with their model.
%
The following listing shows the \FML code of this \textit{virtual model}. Note that the composition relation of Fig.~\ref{fig:BPMNSubsetExample} between \texttt{BPMNProcess} and \texttt{BPMNTask} is realized in the FML code by concept containment. Whenever an instance of \texttt{BPMNTask} is created, it must be in a container (instance of \texttt{BPMNProcess}) and the two instances get linked together. %\noteFabien{A valider}

### 2022-02-23 5-analysis.tex
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

Like in a lot of modeling approaches, these  steps could also be achieved in any order and iteratively. But our goal during the challenge's analysis remains to produce an \FML virtual model architecture.
%for our federation.
As shown in the figure~\ref{fig:MultilevelArchitecture}, we defined two virtual models (Base and Acme processes abstractions) instantiated by two virtual model instances (XSure and Acme processes). As the figure shows the virtual models play the role of metamodels, in the sense of concept definition with a level-agnostic approach. The virtual model instances play the role of models conforming to the metamodel definitions. In the rest of the paper, we use the term metamodel and model to simplify the presentation.
The resulting architecture follows the way the challenge is presented and this
organization allows for the possibility of flexible extensions.

@@

Our analysis of the use case leads to identify two main modeling axes (as presented in the figure \ref{fig:LinguisticAndOntologicInstantiation}):
\begin{itemize}
    \item The horizontal axis is characterized as the ontological instantiation axis, in the sense that the domain type definition (\ie \texttt{ProcessType}) is referenced by an instance definition (\ie \texttt{Process}). In our base metamodel, each domain type definition is referenced by its instance definition, as developed in the section \ref{sec:ProcessEnactment}

\item The vertical axis is viewed as the linguistic instantiation. The Metamodel (a virtual model) can be instantiated as a model thanks to the instantiation mechanism of \FML. This mechanism is similar to the classical object/instance mechanism of the object languages but without any constraint on the referenced virtual model, \ie metamodel. The set of resulting instances comes from several virtual models of any abstraction level as detailed in the section~\ref{sec:AcmeSoftwareDevelopmentProcess}

@@

Based on this approach, we organize our set of models following the architecture of the aforementioned figure \ref{fig:MultilevelArchitecture}. The resulting multi-level architecture is organized in two virtual models, one defining the core concepts metamodel including the definition of Processes and Tasks, and one, the Acme metamodel, extending the previous model to specialize the Acme definitions. The \FML language defines multiple inheritance concept between virtual models as illustrated in the section~\ref{sec:AcmeSoftwareDevelopmentProcess}.

>>>>

Like in a lot of modeling approaches, these  steps could also be achieved in any order and iteratively. But our goal during the challenge's analysis remains to produce an \FML virtual model architecture.
%for our federation.
As shown in the Fig.~\ref{fig:MultilevelArchitecture}, we defined two virtual models (Base and Acme processes abstractions) instantiated by two virtual model instances (XSure and Acme processes). As the figure shows the virtual models play the role of metamodels, in the sense of concept definition with a level-agnostic approach. The virtual model instances play the role of models conforming to the metamodel definitions. In the rest of the paper, we use the term metamodel and model to simplify the presentation.
The resulting architecture follows the way the challenge is presented and this
organization allows for the possibility of flexible extensions.

@@

Our analysis of the use case leads to identify two main modeling axes (as presented in the Fig.~\ref{fig:LinguisticAndOntologicInstantiation}):
\begin{itemize}
    \item The horizontal axis is characterized as the ontological instantiation
      axis, in the sense that the domain type definition (\ie
      \texttt{ProcessType}) is referenced by an instance definition (\ie
      \texttt{Process}). In our base metamodel, each domain type definition is
      referenced by its instance definition, as developed in the Sect.~\ref{sec:ProcessEnactment}

\item The vertical axis is viewed as the linguistic instantiation. The
  Metamodel (a virtual model) can be instantiated as a model thanks to the
  instantiation mechanism of \FML. This mechanism is similar to the classical
  object/instance mechanism of the object languages but without any constraint
  on the referenced virtual model, \ie metamodel. The set of resulting
  instances comes from several virtual models of any abstraction level as
  detailed in the Sect.~\ref{sec:AcmeSoftwareDevelopmentProcess}

@@

Based on this approach, we organize our set of models following the
architecture of the aforementioned Fig.~\ref{fig:MultilevelArchitecture}. The
resulting multi-level architecture is organized in two virtual models, one
defining the core concepts metamodel including the definition of Processes and
Tasks, and one, the Acme metamodel, extending the previous model to specialize
the Acme definitions. The \FML language defines multiple inheritance concept
between virtual models as illustrated in the Sect.~\ref{sec:AcmeSoftwareDevelopmentProcess}.

### 2022-02-23 6-modelpresentation.tex
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

The designed multilevel architecture shown in figure~\ref{fig:MultilevelArchitecture} captures the two use cases described in the Process Challenge.

Note that in all following figures, blue is used to represent \FML virtual models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in figure \ref{fig:mm}), see figures \ref{fig:BPMNSubsetExample}, \ref{fig:MultilevelArchitecture}, \ref{fig:LinguisticAndOntologicInstantiation}, \ref{fig:BaseMetamodel}, \ref{fig:AcmeFullArchitecture},  \ref{fig:ToolingArchitecture}, while brown is used for instances of \textsf{FlexoConcept}, see figures \ref{fig:BPMNSubsetExample}, \ref{fig:MultilevelArchitecture}, \ref{fig:LinguisticAndOntologicInstantiation}, \ref{fig:AcmeFullArchitecture},  \ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9
% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process

@@

In this section, we present the base metamodel presented in Figure~\ref{fig:BaseMetamodel} with the \textit{XSure} insurance domain use case, whose partial description was provided in the challenge description. The requirements \textbf{P1} to \textbf{P19} of XSure insurance domain are straightforwardly implemented by instantiating the \emph{XSure model}, as an instance of base metamodel (the left side of figure~\ref{fig:MultilevelArchitecture}).

Figure~\ref{fig:BaseMetamodel} represents this base metamodel with a \UML-like formalism, well adapted to represent \FML concepts and their instances. A \textit{Concept} of \FML is represented by a \UML class where roles of basic types are attributes. The roles whose types are concepts are represented by an arrow from their concept to their type, with their name on the arrow. The cardinality follows the \UML practice.
For example, the role \texttt{parentActorTypes} of the concept \texttt{ActorType} has the type \texttt{ActorType} and the * cardinality.

Our proposition relies on ontological instantiation as presented in figure \ref{fig:LinguisticAndOntologicInstantiation}, with a common root concept \texttt{ModelingElement}. Two types of ontological instantiation are needed to meet the requirements of the challenge. One provides that some instances conform to only one type (\texttt{Process} and \texttt{Task} do respectively conform to \texttt{ProcessType} and \texttt{TaskType}). While some entities define their conformity to several types (\texttt{Actor} and \texttt{Artifact} do respectively conform to several \texttt{ActorType} and \texttt{ArtifactType}). These instantiations are respectively expressed using the relations \texttt{type} and \texttt{types} between \texttt{Type}, \texttt{Instance} and \texttt{MultiInstance} concepts. In the diagram, the realization of these relations are indicated by a derived relation, \texttt{/type} or \texttt{/types}, as for example between \texttt{Process} and \texttt{ProcessType} or between \texttt{Actor} and \texttt{ActorType}.

@@

We first present the process definition part of the base metamodel, located on the left of Figure \ref{fig:BaseMetamodel}.

@@

\subsubsection{Process enactment}
\label{sec:ProcessEnactment}
We now present the process enactment part of base metamodel, located right of figure~\ref{fig:BaseMetamodel}. All concepts defined in this subsection are either specializations of the \texttt{Instance} concept (if they have exactly one type) or the \texttt{MultiInstance} concept (when they have several types).

@@

The challenge describes in a second part a Software engineering process for a fictional Acme company. The Base metamodel described previously is too generic to capture all domain-specific aspects of this use case. We chose to complete the architectural hierarchy with a specific virtual model, specific to the Acme software development process metamodel, shown in figure~\ref{fig:AcmeFullArchitecture}.
The \textit{Acme} metamodel specializes the Base metamodel, and \textit{Acme model} is an instance of the \textit{Acme} metamodel. The metamodel inheritance is implemented by virtual model's inheritance. Any instance of the Acme model is either an instance of a concept defined in the Base metamodel, or a concept defined in the specialized \textit{Acme} metamodel.
% FML follows a classical object-oriented semantics. ???

@@

The figure~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the architecture. This figure is only partial at instance level and not all instances are represented. It also contains an instance of enacted a \textit{Software Engineering Process} at the bottom. The Acme metamodel specializes some concepts of the Base metamodel, for example, \texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType} extends \texttt{ArtifactType}. These concepts are further specialized  \texttt{CodingTaskType} extends \texttt{SETaskType} and \texttt{CodeArtifactType} extends \texttt{SEArtifactType}.
The Acme metamodel also defines a specialized task type \texttt{Coding} extending \texttt{Task}, and provides the \texttt{Developer} concept extending \texttt{ActorType}. This metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the instances required to capture the challenge use case are defined in the final Acme model (lower part of the figure). Instances are represented with rounded boxes and linguistic instantiation is represented using dashed connectors.

The \textit{Software Development Process} for Acme company is presented in the figure~\ref{fig:AcmeSoftwareDevelopmentProcess}. It is a screen capture of the tool developed for the challenge and described in the next section.

@@

The requirement \textbf{S8} is simply expressed by the definition of the \inst{Testing} instance of \texttt{SETaskType}, \inst{Tester} instance of \texttt{ActorType}, and \inst{Test report} instance of \texttt{ArtifactType}, as shown in the figure \ref{fig:AcmeSoftwareDevelopmentProcess}.

@@

\textbf{S13} requirement has been previously partially fulfilled. This must be completed with an association between an artifact produced and the task that validates it. This is modeled by the reference to \inst{Test Cases} as a produced artifact type in \inst{Test Case design} and the reference to \inst{Test Cases} as a used artifact in \inst{Test Design review}, as shown in figure~\ref{fig:AcmeSoftwareDevelopmentProcess}.

@@

We took advantage of model federation and the availability of diagraming features though the \textit{Diagraming Technology Adapter} to implement two interactive graphical tools built on top of the conceptual levels detailed in previous sections. The figure \ref{fig:ToolingArchitecture} shows the overview of the tooling architecture and is detailed in the next section. A first tool offers a graphical edition of a Process type. The second one offers an enactment feature (the instantiation of a process from its process type), the ability within an enacted process to assign tasks to actors, and a graphical visualization of this process execution.

The Base metamodel is completed with some behaviors implementing an execution semantics for the executed processes. All tasks -- instantiated from \texttt{TaskType} for a given enacted process -- manage a status. This status is either \texttt{Not startable} (when not assigned to a performing actor or when required input artifacts are not available), \texttt{Startable}, \texttt{Started}, \texttt{Completable} (when all output artifacts are ready) or \texttt{Completed}. A \texttt{Task} also manages a set of performing actors, a begin and end date, some used and produced artifacts. The \texttt{Gateway} business logic has also been implemented through the implementation of an abstract \texttt{execute(Process)} behavior. The management of artifacts whose semantics follows rules defined in section~\ref{sec:ProcessEnactment} (\textbf{P13}) is also implemented.

Figure \ref{fig:ToolingArchitecture} details the architecture of these two tools.

@@

The other tool, called \textit{Enacted process graphical editor} and shown on figure~\ref{fig:ScreenshotEnactedProcessEditor}, provides process edition and tasks assignations through a graphical visualization displaying the process being executed. This tool is represented on the right side of the figure~\ref{fig:ToolingArchitecture}. It follows the same architecture pattern of the \textit{ProcessType graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \texttt{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \texttt{TaskType} definition, and required assignments apply (if for example \inst{Coding} \texttt{TaskType} defines COBOL as output programming language, the related task is automatically assigned to \inst{Ann Smith}). Each task has a status as well as a set of performing actors, a begin and end dates, and used and produced artifacts.

>>>>

The designed multilevel architecture shown in Fig.~\ref{fig:MultilevelArchitecture} captures the two use cases described in the Process Challenge.

Note that in all following figures, blue is used to represent \FML virtual models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in Fig.~\ref{fig:mm}), see figures \ref{fig:BPMNSubsetExample}, \ref{fig:MultilevelArchitecture}, \ref{fig:LinguisticAndOntologicInstantiation}, \ref{fig:BaseMetamodel}, \ref{fig:AcmeFullArchitecture},  \ref{fig:ToolingArchitecture}, while brown is used for instances of \textsf{FlexoConcept}, see figures \ref{fig:BPMNSubsetExample}, \ref{fig:MultilevelArchitecture}, \ref{fig:LinguisticAndOntologicInstantiation}, \ref{fig:AcmeFullArchitecture},  \ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9
% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process

@@

In this section, we present the base metamodel presented in Fig.~\ref{fig:BaseMetamodel} with the \textit{XSure} insurance domain use case, whose partial description was provided in the challenge description. The requirements \textbf{P1} to \textbf{P19} of XSure insurance domain are straightforwardly implemented by instantiating the \emph{XSure model}, as an instance of base metamodel (the left side of Fig.~\ref{fig:MultilevelArchitecture}).

Fig.~\ref{fig:BaseMetamodel} represents this base metamodel with a \UML-like formalism, well adapted to represent \FML concepts and their instances. A \textit{Concept} of \FML is represented by a \UML class where roles of basic types are attributes. The roles whose types are concepts are represented by an arrow from their concept to their type, with their name on the arrow. The cardinality follows the \UML practice.
For example, the role \texttt{parentActorTypes} of the concept \texttt{ActorType} has the type \texttt{ActorType} and the * cardinality.

Our proposition relies on ontological instantiation as presented in Fig.~\ref{fig:LinguisticAndOntologicInstantiation}, with a common root concept \texttt{ModelingElement}. Two types of ontological instantiation are needed to meet the requirements of the challenge. One provides that some instances conform to only one type (\texttt{Process} and \texttt{Task} do respectively conform to \texttt{ProcessType} and \texttt{TaskType}). While some entities define their conformity to several types (\texttt{Actor} and \texttt{Artifact} do respectively conform to several \texttt{ActorType} and \texttt{ArtifactType}). These instantiations are respectively expressed using the relations \texttt{type} and \texttt{types} between \texttt{Type}, \texttt{Instance} and \texttt{MultiInstance} concepts. In the diagram, the realization of these relations are indicated by a derived relation, \texttt{/type} or \texttt{/types}, as for example between \texttt{Process} and \texttt{ProcessType} or between \texttt{Actor} and \texttt{ActorType}.

@@

We first present the process definition part of the base metamodel, located on the left of Fig.~\ref{fig:BaseMetamodel}.

@@

\subsubsection{Process enactment}
\label{sec:ProcessEnactment}
We now present the process enactment part of base metamodel, located right of Fig.~\ref{fig:BaseMetamodel}. All concepts defined in this subsection are either specializations of the \texttt{Instance} concept (if they have exactly one type) or the \texttt{MultiInstance} concept (when they have several types).

@@

The challenge describes in a second part a Software engineering process for a fictional Acme company. The Base metamodel described previously is too generic to capture all domain-specific aspects of this use case. We chose to complete the architectural hierarchy with a specific virtual model, specific to the Acme software development process metamodel, shown in Fig.~\ref{fig:AcmeFullArchitecture}.
The \textit{Acme} metamodel specializes the Base metamodel, and \textit{Acme model} is an instance of the \textit{Acme} metamodel. The metamodel inheritance is implemented by virtual model's inheritance. Any instance of the Acme model is either an instance of a concept defined in the Base metamodel, or a concept defined in the specialized \textit{Acme} metamodel.
% FML follows a classical object-oriented semantics. ???

@@

The Fig.~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the architecture. This figure is only partial at instance level and not all instances are represented. It also contains an instance of enacted a \textit{Software Engineering Process} at the bottom. The Acme metamodel specializes some concepts of the Base metamodel, for example, \texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType} extends \texttt{ArtifactType}. These concepts are further specialized  \texttt{CodingTaskType} extends \texttt{SETaskType} and \texttt{CodeArtifactType} extends \texttt{SEArtifactType}.
The Acme metamodel also defines a specialized task type \texttt{Coding} extending \texttt{Task}, and provides the \texttt{Developer} concept extending \texttt{ActorType}. This metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the instances required to capture the challenge use case are defined in the final Acme model (lower part of the figure). Instances are represented with rounded boxes and linguistic instantiation is represented using dashed connectors.

The \textit{Software Development Process} for Acme company is presented in the Fig.~\ref{fig:AcmeSoftwareDevelopmentProcess}. It is a screen capture of the tool developed for the challenge and described in the next section.

@@

The requirement \textbf{S8} is simply expressed by the definition of the \inst{Testing} instance of \texttt{SETaskType}, \inst{Tester} instance of \texttt{ActorType}, and \inst{Test report} instance of \texttt{ArtifactType}, as shown in the Fig.~\ref{fig:AcmeSoftwareDevelopmentProcess}.

@@

\textbf{S13} requirement has been previously partially fulfilled. This must be completed with an association between an artifact produced and the task that validates it. This is modeled by the reference to \inst{Test Cases} as a produced artifact type in \inst{Test Case design} and the reference to \inst{Test Cases} as a used artifact in \inst{Test Design review}, as shown in Fig.~\ref{fig:AcmeSoftwareDevelopmentProcess}.

@@

We took advantage of model federation and the availability of diagraming features though the \textit{Diagraming Technology Adapter} to implement two interactive graphical tools built on top of the conceptual levels detailed in previous sections. The Fig.~\ref{fig:ToolingArchitecture} shows the overview of the tooling architecture and is detailed in the next section. A first tool offers a graphical edition of a Process type. The second one offers an enactment feature (the instantiation of a process from its process type), the ability within an enacted process to assign tasks to actors, and a graphical visualization of this process execution.

The Base metamodel is completed with some behaviors implementing an execution
semantics for the executed processes. All tasks -- instantiated from
\texttt{TaskType} for a given enacted process -- manage a status. This status
is either \texttt{Not startable} (when not assigned to a performing actor or
when required input artifacts are not available), \texttt{Startable},
\texttt{Started}, \texttt{Completable} (when all output artifacts are ready) or
\texttt{Completed}. A \texttt{Task} also manages a set of performing actors, a
begin and end date, some used and produced artifacts. The \texttt{Gateway}
business logic has also been implemented through the implementation of an
abstract \texttt{execute(Process)} behavior. The management of artifacts whose
semantics follows rules defined in Sect.~\ref{sec:ProcessEnactment} (\textbf{P13}) is also implemented.

Fig.~\ref{fig:ToolingArchitecture} details the architecture of these two tools.

@@

The other tool, called \textit{Enacted process graphical editor} and shown on Fig.~\ref{fig:ScreenshotEnactedProcessEditor}, provides process edition and tasks assignations through a graphical visualization displaying the process being executed. This tool is represented on the right side of the Fig.~\ref{fig:ToolingArchitecture}. It follows the same architecture pattern of the \textit{ProcessType graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \texttt{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \texttt{TaskType} definition, and required assignments apply (if for example \inst{Coding} \texttt{TaskType} defines COBOL as output programming language, the related task is automatically assigned to \inst{Ann Smith}). Each task has a status as well as a set of performing actors, a begin and end dates, and used and produced artifacts.

### 2022-02-23 7-requirementssatisfaction.tex
style guidelines: * [sS]ection -> Sect. * [fF]igure -> Fig. * [tT]able -> Tab. * few s/ \ref/~\ref/

In the previous section~\ref{sec:model}, we demonstrated that our solution satisfies all the challenge requirements. In this section, we summarize the different techniques used to satisfy them, without repeating all the justifications. We classify these techniques into three categories:
\begin{itemize}
    \item \emph{syntactic conformance}: the requirements are structurally satisfied when the model conforms to its metamodel,
    \item \emph{constraints checking}: the requirements are fulfilled when all instances satisfy both syntactic conformance and the evaluation of invariants expressed for related concepts,
    \item \emph{tooling}: the requirements are enforced by the implementation of the associated tooling.
\end{itemize}

Table~\ref{tab:RequirementsSatisfactionBaseMetamodel} summarizes the requirement coverage for the base metamodel illustrated by XSure insurance use case while table~\ref{tab:RequirementsSatisfactionAcme} shows requirements satisfaction for Acme software engineering process. Syntactic conformance is more precisely classified into six subcategories:

@@

    \item \emph{Ontological instantiation} (\textit{O.Inst.}): fulfillment of a requirement is obtained by the relation between an instance and its type definition instance (for example the notion of \textit{developer} of \textbf{S3} is implemented by both a concept in \textit{Acme metamodel} and an instance in \textit{Acme model}).

@@

\noteSylvain{updated}

%Table~\ref{tab:RequirementsSatisfactionBaseMetamodel}
%Table~\ref{tab:RequirementsSatisfactionAcme}

>>>>

In the previous Sect.~\ref{sec:model}, we demonstrated that our solution satisfies all the challenge requirements. In this section, we summarize the different techniques used to satisfy them, without repeating all the justifications. We classify these techniques into three categories:
\begin{itemize}
    \item \emph{syntactic conformance}: the requirements are structurally satisfied when the model conforms to its metamodel,
    \item \emph{constraints checking}: the requirements are fulfilled when all instances satisfy both syntactic conformance and the evaluation of invariants expressed for related concepts,
    \item \emph{tooling}: the requirements are enforced by the implementation of the associated tooling.
\end{itemize}

Tab.~\ref{tab:RequirementsSatisfactionBaseMetamodel} summarizes the requirement coverage for the base metamodel illustrated by XSure insurance use case while Tab.~\ref{tab:RequirementsSatisfactionAcme} shows requirements satisfaction for Acme software engineering process. Syntactic conformance is more precisely classified into six subcategories:

@@

    \item \emph{Ontological instantiation} (\textit{O.Inst.}): fulfillment of a requirement is obtained by the relation between an instance and its type definition instance (for example the notion of \textit{developer} of \textbf{S3} is implemented by both a concept in \textit{Acme metamodel} and an instance in \textit{Acme model}, see bottom left of Fig.~\ref{fig:AcmeFullArchitecture}).

@@

%\noteSylvain{updated}

%Tab.~\ref{tab:RequirementsSatisfactionBaseMetamodel}
%Tab.~\ref{tab:RequirementsSatisfactionAcme}
